Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-12-21 journal paper/content/sec-theory.tex
pass through alex's edits

\xhdr{Notation.} Let $\rew_i(n)$ denote the \asedit{agent's realized reward observed by} principal $i$ at local step $n$. \asedit{Put differently, it is the reward collected by algorithm $\alg[i]$ in this local step.} For a global round $t$, let $n_i(t)$ denote the number of global rounds before $t$ in which principal $i$ is chosen. Note that
    \[ \PMR_i(t):=  \E\sbr{ r_t\mid i_t = i }
        = \E\sbr{ \rew_i(n_i(t)+1) }. \]

@@

\xhdr{MAB algorithms.}
\asedit{We consider two baseline algorithms. The main one,} called  \DynGreedy, chooses an arm $a$ with the largest posterior mean reward $\E[\mu_a \mid \cdot ]$ given all information currently available to the algorithm. \asedit{A more primitive baseline,} called \StaticGreedy, chooses an arm $a$ with the largest prior mean reward $\E[\mu_a]$, and uses this arm in all rounds.

@@

\begin{definition}
\asedit{One MAB algorithm \emph{deviates} from another} at (local) step $n$ if there is a set $H$ of histories over the previous local steps such that both algorithms lead to $H$ with positive probability, and choose different distributions over arms given any history $h\in H$. \asedit{We say $\alg$ deviates from $\alg'$} \emph{starting from} step $n_0$ if $n=n_0$ is the smallest step \asedit{when this happens.}
\end{definition}

@@

\asedit{The sudden-death property in Lemma~\ref{lm:DG-sudden} holds because principal $1$ appears at least as good or better to the next agent (by the Bayesian-monotonicity property), whereas principal $2$ appears the same as before. Principal $2$ needs new data in order to improve, and it does not receive new data unless the response function is randomized.}

@@

\asedit{Note that we make no positive prediction if one MAB algorithm is only slightly better than the other. Next we show that such predictions are essentially impossible.}

@@

\asedit{We argue below that \alg[1] can be a better algorithm than \alg[2]. This is consistent with Theorem~\ref{thm:random-clean}, since the BIR-dominance condition does not hold.}

@@

\asedit{We also use \refeq{eq:BIR-modification} to argue that} \alg[2] is \bmonotone. This follows from \eqref{eq:BIR-modification} since both \alg[1] and \alggr are \bmonotone. The latter follows from a ``monotonicity-in-information" property of the ``greedy step": it can only get better with more information, see Lemma~\ref{lm:MII}.

@@

Recall that by Theorem~\ref{thm:random-clean} \DynGreedy cannot be played in equilibrium when it is BIR-dominated by some feasible, $T$-oblivious algorithm and the time horizon $T$ is large enough. \asedit{Also recall that pure and mixed Nash equilibria are the same in our model.}

\asedit{Corollary~\ref{cor:random-greedy} holds more generally: it suffices to assume that feasible algorithms are \bmonotone and closed under the greedy modification. The latter means that the greedy modification of any feasible algorithm is also feasible, for any mixing parameter $p$.}

@@

We frame our contributions in terms of the relationship between \competitiveness (as expressed by the ``hardness" of the agents' response function $\respF$),
 %and \rationality on one side,
 and adoption of better algorithms.
\asdelete{Recall that both \competitiveness (of the game between the two principals) and \rationality (of the agents) are controlled by the response function $\respF$.}

@@

\xhdr{Secondary story.}
Let us zoom in on the symmetric  \HardMaxRandom model. Competitiveness within this model are controlled by the baseline probability $\eps_0 = \respF(- 1)$, which goes smoothly between the two extremes of \HardMax ($\eps_0=0$, tough competition) and the uniform choice ($\eps_0=\tfrac12$, no competition). 
\asedit{Recall} % For clarity, we assume 
that principal's utility is the number of agents.

>>>>

\xhdr{Notation.} Let $\rew_i(n)$ denote the agent's realized reward observed by principal $i$ at local step $n$. Put differently, it is the reward collected by algorithm $\alg[i]$ in this local step. For a global round $t$, let $n_i(t)$ denote the number of global rounds before $t$ in which principal $i$ is chosen. Note that
    \[ \PMR_i(t):=  \E\sbr{ r_t\mid i_t = i }
        = \E\sbr{ \rew_i(n_i(t)+1) }. \]

@@

\xhdr{MAB algorithms.}
We consider two baseline algorithms. The main one, called  \DynGreedy, chooses an arm $a$ with the largest posterior mean reward $\E[\mu_a \mid \cdot ]$ given all information currently available to the algorithm. A more primitive baseline, called \StaticGreedy, chooses an arm $a$ with the largest prior mean reward $\E[\mu_a]$, and uses this arm in all rounds.

@@

\begin{definition}
One MAB algorithm \emph{deviates} from another at (local) step $n$ if there is a set $H$ of histories over the previous local steps such that both algorithms lead to $H$ with positive probability, and choose different distributions over arms given any history $h\in H$. We say $\alg$ deviates from $\alg'$ \emph{starting from} step $n_0$ if $n=n_0$ is the smallest step when this happens.
\end{definition}

@@

The sudden-death property in Lemma~\ref{lm:DG-sudden} holds because principal $1$ appears at least as good or better to the next agent (by the Bayesian-monotonicity property), whereas principal $2$ appears the same as before. Principal $2$ needs new data in order to improve, and it does not receive new data unless the response function is randomized.

@@

Note that we make no positive prediction if one MAB algorithm is only slightly better than the other. Next we show that such predictions are essentially impossible.

@@

We argue below that \alg[1] can be a better algorithm than \alg[2]. This is consistent with Theorem~\ref{thm:random-clean}, since the BIR-dominance condition does not hold.

@@

We also use \refeq{eq:BIR-modification} to argue that \alg[2] is \bmonotone. This follows from \eqref{eq:BIR-modification} since both \alg[1] and \alggr are \bmonotone. The latter follows from a ``monotonicity-in-information" property of the ``greedy step": it can only get better with more information, see Lemma~\ref{lm:MII}.

@@

Recall that by Theorem~\ref{thm:random-clean} \DynGreedy cannot be played in equilibrium when it is BIR-dominated by some feasible, $T$-oblivious algorithm and the time horizon $T$ is large enough. Also recall that pure and mixed Nash equilibria are the same in our model.

Corollary~\ref{cor:random-greedy} holds more generally: it suffices to assume that feasible algorithms are \bmonotone and closed under the greedy modification. The latter means that the greedy modification of any feasible algorithm is also feasible, for any mixing parameter $p$.

@@

We frame our contributions in terms of the relationship between \competitiveness (as expressed by the ``hardness" of the agents' response function $\respF$),
 %and \rationality on one side,
 and adoption of better algorithms.

@@

\xhdr{Secondary story.}
Let us zoom in on the symmetric  \HardMaxRandom model. Competitiveness within this model are controlled by the baseline probability $\eps_0 = \respF(- 1)$, which goes smoothly between the two extremes of \HardMax ($\eps_0=0$, tough competition) and the uniform choice ($\eps_0=\tfrac12$, no competition). 
Recall % For clarity, we assume 
that principal's utility is the number of agents.

### 2020-12-21 journal paper/content/sim_details.tex
pass through alex's edits

\xhdr{Challenges.} An ``atomic experiment" is a competition game between a given pair of bandit algorithms, in a given competition model, on a given 
%instance of a 
multi-armed bandit problem (and each such experiment is run many times to reduce variance). Accordingly, we have a three-dimensional space of atomic experiments one needs to run and interpret: \{pairs of algorithms\} x \{competition models\} x \{bandit \asedit{problems}\}, and we are looking for findings that are consistent across this entire space. It is essential to keep each of the three dimensions small yet representative. In particular, we need to capture a huge variety of bandit algorithms and bandit instances with only a few representative examples. Further, we need a succinct and informative summarization of results within one atomic experiment and across multiple experiments (\eg see Table~\ref{fig:market_share}).

@@

We look for qualitative differences between these three classes under competition. We take a representative algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other representative algorithms are chosen.
We use \Thompson (\TS) from the ``adaptive-exploration" algorithms, \DynamicEpsGreedy (\DEG) from the ``exploration-separating" algorithms,%
\footnote{In each round, \DynamicEpsGreedy algorithm explores uniformly with a predetermined probability $\eps$, and ``exploits" with the remaining probability, choosing an arm with maximal posterior mean reward given the current data. We use $\eps=5\%$ throughout. Our pilot experiments show that choosing a different $\eps$ does not qualitatively change the results.}  and \DynGreedy (\DG) algorithm  as in Section~\ref{sec:theory-prelims}. For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.

@@

\xhdr{MAB instances.}
We consider \asedit{bandit problems} with $K=10$ arms and Bernoulli rewards. \asedit{The \emph{\MRV} $(\mu(a):\; a\in A)$ is initially} drawn from some distribution, termed \emph{MAB instance}. We consider three MAB instances:
\begin{enumerate}
\item \emph{Needle-In-Haystack}: one arm (the ``needle") is chosen uniformly at random. This arm has mean reward $.7$, and the remaining ones have mean reward $.5$.

>>>>

\xhdr{Challenges.} An ``atomic experiment" is a competition game between a given pair of bandit algorithms, in a given competition model, on a given 
%instance of a 
multi-armed bandit problem (and each such experiment is run many times to reduce variance). Accordingly, we have a three-dimensional space of atomic experiments one needs to run and interpret: \{pairs of algorithms\} x \{competition models\} x \{bandit problems\}, and we are looking for findings that are consistent across this entire space. It is essential to keep each of the three dimensions small yet representative. In particular, we need to capture a huge variety of bandit algorithms and bandit instances with only a few representative examples. Further, we need a succinct and informative summarization of results within one atomic experiment and across multiple experiments (\eg see Table~\ref{fig:market_share}).

@@

We look for qualitative differences between these three classes under competition. We take a representative algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other representative algorithms are chosen.
We use \Thompson (\TS) from the ``adaptive-exploration" algorithms\footnote{\gaedit{We provide details on \Thompson in Appendix~\ref{app:bg}.}} \DynamicEpsGreedy (\DEG) from the ``exploration-separating" algorithms,%
\footnote{\gaedit{\DynamicEpsGreedy proceeds as follows. In each round, with probability $\eps$ it uniformly at random selects an arm to pull from the full set of arms and} ``exploits" with the remaining probability, choosing an arm with maximal posterior mean reward given the current data. We use $\eps=5\%$ throughout. Our pilot experiments show that choosing a different $\eps$ does not qualitatively change the results.}  and \DynGreedy (\DG) algorithm  as in Section~\ref{sec:theory-prelims}. For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.

@@

\xhdr{MAB instances.}
We consider bandit problems with $K=10$ arms and Bernoulli rewards. The \emph{\MRV} $(\mu(a):\; a\in A)$ is initially drawn from some distribution, termed \emph{MAB instance}. We consider three MAB instances:
\begin{enumerate}
\item \emph{Needle-In-Haystack}: one arm (the ``needle") is chosen uniformly at random. This arm has mean reward $.7$, and the remaining ones have mean reward $.5$.

### 2020-12-21 journal paper/journal_main.tex
A pass addressing YK comments.
pass through alex's edits

\title{\vspace{-12mm}Competing Bandits:\\
The Perils of Exploration under Competition%
\thanks{This is a merged and final version of two conference publications,
\citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19},
with a unified and streamlined presentation and expanded related work and background materials. All theoretical results are from \citet{CompetingBandits-itcs18}, and all numerical simulations are from \citet{CompetingBandits-ec19}. Appendices~\ref{app:bg},\ref{app:examples} are completely new compared to the conference versions.\newline \indent 
We would like to thank Ian Ball, Yeon-Koo Che, Glen Weyl and Sven Rady for helpful comments and conversations. Likewise, we thank the audience of the conference and seminar talks on this work.
%seminar participants at Columbia and conference participants at Innovations in Theoretical Computer Science 2018, ACM Economics and Computation 2019 and the MIT Conference on Digital Experimentation 2020, for helpful comments and conversations.
All errors are our own.}}
% \textit{Competing Bandits: Learning under Competition}, at Innovations in Theoretical Computer Science 2018, and \textit{The Perils of Exploration under Competition: A Computational Modeling Approach}, at ACM Economics and Computation 2019.}}

@@

\vspace{-5mm}
\begin{abstract}
Most online platforms strive to learn from interactions with users, and many engage in \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. We \asedit{study} the interplay between \emph{exploration} and \emph{competition}: how such platforms balance the exploration for learning and the competition for users. Here users play three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents which choose among the competing platforms.

We consider a stylized duopoly model in which two firms face \asedit{the same multi-armed bandit problem}. Users arrive one by one and choose between the two firms, so that each firm makes progress on its bandit instance only if it is chosen. Through a mix of theoretical results and numerical simulations, we study whether and to what extent competition incentivizes the adoption of better bandit algorithms, and whether it leads to welfare increases for users. We find that stark competition induces firms to commit to a ``greedy" bandit algorithm that leads to low welfare. However, weakening competition by providing firms with some ``free" users incentivizes better exploration strategies and increases welfare.  We investigate two channels for weakening the competition: relaxing the rationality of users and giving one firm a first-mover advantage. Our findings are closely related to the ``competition vs. innovation" relationship, and elucidate the first-mover advantage in the digital economy.

@@

\section{Related work}
\label{sec:related-work}
\subfile{content/related_work}

@@

%\newpage
\section{Theoretical results: the \TheoryModel}
\label{sec:theory}
\subfile{content/sec-theory}

>>>>

\title{\vspace{-12mm}Competing Bandits:\\
The Perils of Exploration under Competition%
\thanks{This is a merged and final version of two conference publications,
\citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19},
with a unified and streamlined presentation and expanded related work and background materials. All theoretical results are from \citet{CompetingBandits-itcs18}, and all numerical simulations are from \citet{CompetingBandits-ec19}. Appendices~\ref{app:bg},\ref{app:examples} are completely new compared to the conference versions.\newline \indent 
We would like to thank Ian Ball, Yeon-Koo Che, Sara Shahanaghi, Glen Weyl and Sven Rady for helpful comments and conversations. Likewise, we thank the audience of the conference and seminar talks on this work.
%seminar participants at Columbia and conference participants at Innovations in Theoretical Computer Science 2018, ACM Economics and Computation 2019 and the MIT Conference on Digital Experimentation 2020, for helpful comments and conversations.
All errors are our own.}}
% \textit{Competing Bandits: Learning under Competition}, at Innovations in Theoretical Computer Science 2018, and \textit{The Perils of Exploration under Competition: A Computational Modeling Approach}, at ACM Economics and Computation 2019.}}

@@

\vspace{-5mm}
\begin{abstract}
Most online platforms strive to learn from interactions with users, and many engage in \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. We study the interplay between \emph{exploration} and \emph{competition}: how such platforms balance the exploration for learning and the competition for users. Here users play three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents which choose among the competing platforms.

We consider a stylized duopoly model in which two firms face the same multi-armed bandit problem. Users arrive one by one and choose between the two firms, so that each firm makes progress on its bandit problem only if it is chosen. Through a mix of theoretical results and numerical simulations, we study whether and to what extent competition incentivizes the adoption of better bandit algorithms, and whether it leads to welfare increases for users. We find that stark competition induces firms to commit to a ``greedy" bandit algorithm that leads to low welfare. However, weakening competition by providing firms with some ``free" users incentivizes better exploration strategies and increases welfare.  We investigate two channels for weakening the competition: relaxing the rationality of users and giving one firm a first-mover advantage. Our findings are closely related to the ``competition vs. innovation" relationship, and elucidate the first-mover advantage in the digital economy.

@@

\newpage
\section{Related work}
\label{sec:related-work}
\subfile{content/related_work}

@@

\section{Theoretical results: the \TheoryModel}
\label{sec:theory}
\subfile{content/sec-theory}
